Inferring the existence and direction of causal associations in the face of measurement error

### 2016-12-28 docs/manuscript.Rmd
up;dated|
updated

---
title: "Determining the existence and direction of causality in the face of measurement error"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    keep_tex: true
bibliography: manuscript.bib
csl: nature.csl
---

@@

@@

@@

fig_nums(
	"d_relationship_figure", 
	caption="Plots depicting the parameter space in which the function $d = cor(x, x_o) - cor(x,y)cor(y, y_o)$ is negative. When $d$ is negative the Steiger test is liable to infer the wrong direction of causality. $d$ is plotted on the y axis, with measurement precision of $x$ ($cor(x,x_o)$) on the x-axis. Higher measurement precision of $y$ ($cor(y,y_o)$) is depicted by lighter coloured lines, and columns of boxes are for different true values of $cor(x,y)$. Shaded regions show the parameter space where $d$ is negative. The graph shows that for the majority of the parameter space of the function, $d$ is positive, especially where causal relationships are relatively weak.",
	display=FALSE
)

fig_nums(
	"causality_exists_tpr",
	caption="Outcomes were simulated to be causally influenced by exposures with varying degrees of measurement imprecision applied to the exposure variable (x axis). True positive rates (y axis) for MR and CIT were compared with for varying levels of measurement imprecision in the outcome variable (rows of boxes) and sample sizes (columns of boxes).",
	display=FALSE
)

@@

@@

```
```{r read_data }

@@

With our ability to characterize the human phenome ever improving it is becoming increasingly common to use statistical tests to infer the causal relationships between correlated variables. A simple and commonly used method for inferring if an exposure is on the causal pathway to an outcome is through using genetic instruments in mediation analysis, where the effect of an instrument on the outcome is tested before and after adjusting for the exposure. We show that in the face of measurement error this can lead to erroneous results, and that increasing sample size, rather than reducing bias, can often have the effect of increasing certainty in the wrong answer. Commonly used tools that are predicated on this methodology such as the causal inference test (CIT) are shown to be susceptible. We demonstrate that Mendelian randomization (MR) is a method for causal inference that is robust to measurement error, and that a simple extension to MR can provide a method to estimate the direction of causality between correlated variables even when the biology of the instrument is unknown. Using this method we demonstrate that for correlations between DNA methylation and gene expression levels, DNA methylation is likely to be the causal factor. We show that using MR to infer causal directions is straightforward, but because measurement error is ubiquitous in observational data, mediation-based causal inference in particular should be treated with caution.

@@

The last decade has seem a rapid acceleration in technological advances that enable us to measure the human phenome in ever increasing detail, furnishing biologists with the tantalising opportunity to map the molecular basis of complex traits. This does, however, require overcoming the statistical challenge of correctly defining the causal architectures that underlie these high dimensional data. For example, identifying metabolic perturbations that lead to disease [@Kettunen2012], or understanding the influence of epigenomic influences on gene expression [@Bird2002] all depend on understanding the causal nature of any putative observational associations. Despite the ubiquity of the question, commonly used observational study data are notoriously susceptible to pitfalls in making causal inference, with simple regression-based techniques unable to distinguish a true causal association from reverse causation or confounding [@DaveySmith2004].

An established statistical design that can resolve this problem is randomisation. Given the hypothesis that an exposure is causally related to an outcome, randomisation can be employed to assess the causal nature of the association by randomly splitting the sample into two groups, subjecting one group to the exposure and leaving the other as a control. The association between the exposure and the outcome in this setting provides a robust estimate of the causal relationship. This is the theoretical basis behind randomised control trials, but in practice randomisation is often impossible to implement in an experimental context due to cost, scale or inability to manipulate the exposure. The principle, however, can be employed in extant observational data through the use of genetic variants associated with the exposure (instruments), where the inheritance of an allele serves as a random lifetime allocation of differential exposure levels [@DaveySmith2004]. Two statistical approaches to exploiting the properties of genetic instruments are widely used: mediation and Mendelian randomisation (MR).

Mediation is a commonly used statistical method for making causal inference in observational data, and it exists in many forms, from regression based systems  [@Baron1986] to structural equation modeling (SEM) [@Aten2008]. In its simplest form mediation analysis seeks to explain the mechanism that underlies an observed relationship between an independent and a dependent variable by introducing a third, 'mediator' variable. If the association between the independent and dependent variables disappears when conditioning on the mediator variable then this provides evidence that the independent variable influences the dependent variable through the mediator.

This simple form, however, is prone to being unreliable. Alternative models of reverse causation or confounding are often indistinguishable from the hypothesised model of the independent variable causing the dependent variable via the mediator [@Fewell2007]. Introducing pseudo-randomisation by using a genetic instrument as the independent variable can be used to improve on these issues. Supposing an exposure has a causal effect on an outcome, then an ‘instrument’ that causes the exposure (a single nucleotide polymorophism (SNP), for example) should also influence the outcome. Therefore the influence of the instrument on the outcome conditional on the exposure should be zero. This forms the basis of a number of methods such as genetical genomics [@Schadt2005], the regression-based causal inference test (CIT) [@Millstein2009; @Millstein2016], an SEM implementation in the NEO software [@Aten2008], and various other methods including Bayesian approaches [@Waszak2015]. They have been employed by a number of recent publications that make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015; @Yuan2014; @Tang2014; @Hong2015].

MR employs the same data under but uses a different approach in how the instrumental variable is employed. Here an instrumental variable (typically a SNP) that has a robost, direct association with the exposure is used as a surrogate for the exposure in an association test with the outcome, avoiding the problems of confounding or reverse causation by testing for association between the outcome and the SNP. The causal association between the exposure and the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure.

By utilising genetic instruments in different ways, mediation and MR models have properties that confer some advantages and some disadvantages. The main advantage of mediation over MR is that the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure. In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables by performing the test in both directions and choosing the model with the strongest evidence.

Such is not the case for MR, where in order to infer the direction of causality the instrument must be known to influence the exposure directly, associating with the outcome only through the exposure. This requirement can be problematic because if there exists a putative association between two variables, with the instrumental SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the direct effect of the SNP (i.e., for which of the two variables is the SNP a valid instrument?). By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, so if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure. Such a situation can arise in many scenarios. Genome wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015]. For example, the same FTO variant that is associated with body mass index (BMI) is also associated with C-reactive protein (CRP). It is known that this is because BMI causes CRP [@Timpson2011], but supposing we only had *a priori* knowledge that FTO associated with CRP, and we used it to instrument CRP in an MR analysis of CRP against BMI, we would mistakenly infer that CRP causes BMI. Another situation is where the causal directions between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Shakhbazov2016]?

@@

Measurement (or observational) error is the difference between the measured value of a quantity and its true value. Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009]. Array technology is now commonly used to obtain high throughput phenotyping at low cost, but they come with the problem of having imperfect resolution, for instance methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value [@Harper2013; @Chen2013a]. Relatedly, if the measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].

Given the near-ubiquitous presence of measurement error in biological data it is vital to understand its impact on the tools we use for causal inference. Here we show using theory and simulations how non-differential measurement error and measurement imprecision can lead to unreliable causal inference in the mediation-based CIT method. We also present an extension to MR that allows researchers to ascertain the causal direction even when the biology of the instrument is not fully understood. This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially, and we apply this new method to infer the direction of causation between DNA methylation levels and gene expression levels.

@@

We used the Steiger test to test if DNA methylation levels are more likely to causally influence gene expression levels or vice versa. To do this we obtained a list of `r nrow(qtl)` gene expression - DNA methylation associations as reported in Shakhbazov et al [-@Shakhbazov2016]. These were filtered to be located on the same chromosome, have robost correlations after correcting for multiple testing, and to share a SNP that had a robust cis-acting effect on both the DNA methylation probe and the gene expression probe. Using the reported summary data we used the Steiger test of two independent correlations to identify which direction of causality was most likely for each association. We found that the causal direction commonly goes in both directions (`r ctab("shakhtab")`) but methylation levels were more likely to influence expression levels than vice versa (probability of expression causing methylation = `r shakhtest3$p.value` (95% CI `r shakhtest3$conf.int[1]` - `r shakhtest3$conf.int[2]`) amongst `r sum(shakhbazov$dir_p < 0.05)` associations with $p_{Steiger} < 0.05$).

@@

## Tables

```{r shakhtab }

`r tab_nums("shakhtab")`

@@

```{r d_relationship_figure, fig.width=10, fig.height=6 }

@@

`r fig_nums("d_relationship_figure")`

@@

@@

where $\hat{x} = \hat{\alpha}_1 + \hat{\beta}_1 g$. Writing in scale free terms, $\rho_{g, x}$ denotes the correlation between $g$ and the exposure variable $x$, and it is expected that $\rho_{g, x} > \rho_{g, y}$ because $\rho_{g, y} = \rho_{g, x}\rho_{x, y}$, where $\rho_{x, y}$ is the causal association between $x$ and $y$ (which is likely to be less than 1). In the presence of measurement error in $x$ and $y$, however, Steiger test will instead be assessing the inequality $\rho_{g, x_o} > \rho_{g, y_o}$, which can be simplified:

$$
\begin{aligned}
\rho_{g, x_o} & > \rho_{g, y_o} \\
\rho_{g, x} \rho_{x, x_o} & > \rho_{g,y}\rho_{y,y_o}\\
\rho_{g, x} \rho_{x, x_o} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_o}\\
\rho_{x, x_o} & > \rho_{x,y}\rho_{y,y_o}
\end{aligned}
$$

## Appendix 3

The Steiger test is susceptible to inferring the wrong direction of causality under some measurement error parameters. For an observed association between the instrument $g$ and the measured exposure $x_o$, $\rho_{g,x_o}$, the association between $g$ and the true exposure (without measurement error) $x$, $\rho_{g,x} = \frac{\rho_{x,x_o}}{\rho_{g,x_o}}$ where the bounds $\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1$ are the theoretical limits of the correlation between true and measured $x$. The same logic applies to the association between the instrument and the outcome variable.

For the values $\rho_{g,x_o} = 0.3$ and $\rho_{g,y_o} = 0.05$, the possible true $\rho_{g,x}$ and $\rho_{g,y}$ enumerated for all possible values of measurement error can be visualised as so:

```{r, fig.width=6, fig.height=6}

sensitivity_example <- steiger_sensitivity(0.3, 0.05)
sensitivity_example$pl

Here, where $z = \rho_{g,y} - \rho_{g,x} < 0$ the measurement error range in $x_o$ and $y_o$ will not change the qualitative inference of the direction of causality. To evaluate the sensitivity, $S$ of the Steiger test to measurement error, the objective is to calculate the proportion of the total absolute volume of this function that is above the $z = 0$ plane,

@@

$V_{z}$, the total volume of the function, can be obtained analytically by solving the double integral:

@@

$S$ can take values from 0 to 0.5, where low values suggest low sensitivity of the Steiger test to measurement error. For the values used to plot the function above, $\rho_{g,x_o} = 0.3$ and $\rho_{g,y_o} = 0.05$, we obtain $S = `r round(sensitivity_example$sensitivity,3)`$. Hence, the majority of the parameter space of measurement error will not change the inferred causal direction of $x \rightarrow y$ for these values.

>>>>

---
title: "Inferring the existence and direction of causal associations in the face of measurement error"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    fig_caption: no
    keep_tex: true
bibliography: manuscript.bib
csl: nature.csl
---

@@

```{r read_data }

```{r shakhbazov_tests }

```{r steiger_sensitivity }

@@

fig_nums(
	"dag_figure",
	caption="Gene expression levels (blue blocks) and DNA methylation levels (green triangles) may be correlated but the causal direction is unknown. If a SNP (yellow circle) is associated with both DNA methylation and gene expression levels then it can be used as an instrument, but there are two competing models for these variables. a) Methylation causes gene expression. The left figure shows that the SNP influences methylation levels that in turn influence gene expression levels. The right figure shows the directed acyclic graph that represents this model. Faded symbols represent the measured values whereas solid symbols represent the true values. b) The same as in A, except the causal direction is from gene expression to DNA methylation.",
	display=FALSE
)

@@

fig_nums(
	"steiger_sensitivity_plot", 
	caption="a) We can predict the values the Steiger test would take (z-axis) for different potential values of measurement error (x and y axes), drawn here as the blue surface. When $\\rho_{g,y} > \\rho_{g,x}$, as denoted by range of values where the blue surface is above the red plane, those values of measurement error lead to our observed Steiger test inferring the wrong causal direction. Where the blue surface lies below the red plane, these measurement error values support the inferred causal direction of X to Y. A measure of sensitivity, therefore, is the volume of the positive space bound by the blue surface as a proportion of the total space bound by the blue surface. Sensitivity of 0 denotes no measurement error values that could lead to inferrence of incorrect direction, whereas the maximum value of 0.5 denotes measurement error equal chance of either direction across all measurement error ranges. In this case, where $\\rho_{g,x}^2 = 0.01$ and $\\rho_{x,y}^2 = 0.1$, the sensitivity is 0.18. b) The influence of $\\rho_{x,y}$ (x-axis) on the Steiger sensitivity values (y-axis) for different values of $\\rho_{g,x}^2$. Inferring the direction of causality is more susceptible to measurement error when there are stronger causal associations between X and Y.",
	display=FALSE
)

fig_nums(
	"causality_exists_tpr",
	caption="Outcomes were simulated to be causally influenced by exposures with varying degrees of measurement imprecision applied to the exposure variable (x axis). True positive rates (y axis) for MR and CIT were compared for varying levels of measurement imprecision in the outcome variable (rows of boxes) and sample sizes (columns of boxes).",
	display=FALSE
)

@@

fig_nums(
	"shakhplot",
	caption=paste0("Using ", nrow(qtl_orig), " putative associations between DNA methylation and gene expression we used the Steiger test to infer the direction of causality between them. The rightmost bar shows the proportion of associations for each of the two possible causal directions (colour key) assuming no measurement error in either gene expression or DNA methylation levels. The proportions change when we assume different levels of measurement error in gene expression levels (x-axis) or DNA methylation levels (columns of boxes). If there is systematically higher measurement error in one platform it will appear to be less likely to be the causal factor."),
	display=FALSE
)

@@

@@

With our ability to characterize the human phenome ever improving it is becoming increasingly common to use statistical tests to infer the causal relationships between correlated variables. A simple and commonly used method for inferring if an exposure is on the causal pathway to an outcome is through using genetic instruments in mediation analysis, where the effect of an instrument on the outcome is tested before and after adjusting for the exposure. We show that in the face of measurement error this can lead to erroneous inference of both existence of causality and the causal direction, and that increasing sample size, rather than reducing bias, can often have the effect of increasing certainty in the wrong answer. Commonly used tools that are predicated on this methodology such as the causal inference test (CIT) are shown to be susceptible. We demonstrate that Mendelian randomization (MR) is a method for causal inference that is robust to measurement error, and introduce a simple extension to estimate the direction of causality between correlated variables even when the biology of the instrument is unknown. This extension is less susceptible to measurement error than mediation-based analyses. Using these methods we demonstrate empirically that the inferred direction of causality between correleted DNA methylation and gene expression levels is highly susceptible to differences in measurement error between the two assaying platforms, demanding caution when making such claims about causality.

The last decade has seem a rapid acceleration in technological advances that enable us to assess the human phenome in ever increasing detail, furnishing biologists with the tantalising opportunity to map the molecular basis of complex traits. This does, however, require overcoming the statistical challenge of correctly defining the causal architectures that underlie these high dimensional data. For example, identifying metabolic perturbations that lead to disease [@Kettunen2012], or understanding the influence of epigenomic influences on gene expression [@Bird2002] all depend on understanding the causal nature of any putative observational associations.

@@

Widely available observational study data are notoriously susceptible to pitfalls in making causal inference, with basic regression-based techniques unable to distinguish a true causal association from reverse causation or confounding [@DaveySmith2004; @Phillips1991; @DaveySmith2002]. In response to this, genetic instrumentation has emerged as a valuable technique for improving the reliability of causal inference in observational data, and with the coincident rise in genome-wide association studies, it is now a prominent tool that is applied in several different guises, and has been applied many forms of high throughput 'omic variables. However, an often neglected pitfall to these approaches is the influence of non-differential measurement error on the reliability of these methods.

Non-differential measurement error (or observational error, henceforth referred to simply as measurement error) is the difference between the measured value of a quantity and its true value. Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009]. Array technology is now commonly used to obtain high throughput phenotyping at low cost, but they come with the problem of having imperfect resolution, for instance methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value [@Harper2013; @Chen2013a]. Relatedly, if the measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].

Given the near-ubiquitous presence of measurement error in phenomic data it is vital to understand its impact on the tools we use for causal inference. An established study design that can provide information about causality is randomisation. Given the hypothesis that trait A (henceforth referred to as the exposure) is causally related to trait B (henceforth referred to as the outcome), randomisation can be employed to assess the causal nature of the association by randomly splitting the sample into two groups, subjecting one group to the exposure and leaving the other as a control. The association between the exposure and the outcome in this setting provides a robust estimate of the causal relationship. This provides the theoretical basis behind randomised control trials, but in practice randomisation is often impossible to implement in an experimental context due to cost, scale or inability to manipulate the exposure. The principle, however, can be employed in extant observational data through the use of genetic variants associated with the exposure (instruments), where the inheritance of an allele serves as a random lifetime allocation of differential exposure levels [@DaveySmith2004]. Two statistical approaches to exploiting the properties of genetic instruments are widely used: mediation-based approaches and Mendelian randomisation (MR).

Mediation-based approaches employ genetic instruments (typically single nucleotide polymorphisms, SNPs) to orient the causal direction between the exposure and the outcome. If a SNP is associated with an exposure, and the exposure is associated with some outcome, then it logically follows that the implied influence of the SNP on the outcome will disappear when conditioning on the exposure. Here, the exposure 'mediates' the association between the SNP and the outcome, providing information about the causal influence of the exposure on the outcome. This forms the basis of a number of methods such as genetical genomics [@Schadt2005], the regression-based causal inference test (CIT) [@Millstein2009; @Millstein2016], an SEM implementation in the NEO software [@Aten2008], and various other methods including Bayesian approaches [@Waszak2015]. They have been employed by a number of recent publications that make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015; @Yuan2014; @Tang2014; @Hong2015].

MR can operate on the same data - phenotypic measures of the exposure and the outcome variables and a genetic instrument for the exposure - but the genetic instrument is employed in a subtly different manner. Here the SNP, which is assumed to have a direct association with the exposure, is used as a surrogate for the exposure. The causal effect of the exposure and the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure.

By utilising genetic instruments in different ways, mediation-based analysis and MR models have properties that confer some advantages and some disadvantages. The main advantage of mediation over MR is that the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure. In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables by performing the test in both directions and choosing the model with the strongest evidence.

Such is not the case for MR, where in order to infer the direction of causality the instrument must be known to influence the exposure directly, associating with the outcome only through the exposure. This requirement can be problematic because if there exists a putative association between two variables, with the instrumental SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the direct effect of the SNP (i.e., for which of the two variables is the SNP a valid instrument? `r cfig("dag_figure")`). By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, so if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure. Such a situation can arise in many scenarios. Genome-wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015]. For example, the same FTO variant that is associated with body mass index (BMI) is also associated with C-reactive protein (CRP). It is known that this is because BMI causes CRP [@Timpson2011], but supposing we only had *a priori* knowledge that FTO associated with CRP, and we used it to instrument CRP in an MR analysis of CRP against BMI, we would mistakenly infer that CRP causes BMI. Another situation is where the causal directions between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Shakhbazov2016] (`r cfig("dag_figure")`)?

@@

Here we show using theory and simulations how non-differential measurement error and measurement imprecision can lead to unreliable causal inference in the mediation-based CIT method. We also present an extension to MR that allows researchers to ascertain the causal direction even when the biology of the instrument is not fully understood. This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially, and we apply this new method to infer the direction of causation between DNA methylation levels and gene expression levels.

@@

We used the Steiger test to infer the direction of causality between DNA methylation levels and gene expression levels. To do this we obtained a list of `r nrow(qtl_orig)` gene expression - DNA methylation associations as reported in Shakhbazov et al [-@Shakhbazov2016]. These were filtered to be located on the same chromosome, have robost correlations after correcting for multiple testing, and to share a SNP that had a robust cis-acting effect on both the DNA methylation probe and the gene expression probe. Using the reported summary data we used the Steiger test of two independent correlations to identify which direction of causality was most likely for each association. We found that the causal direction commonly goes in both directions (`r ctab("shakhtab")`) but methylation levels were more likely to influence expression levels than vice versa (probability of expression causing methylation = `r shakhtest3$p.value` (95% CI `r shakhtest3$conf.int[1]` - `r shakhtest3$conf.int[2]`) amongst `r sum(shakhbazov$dir_p < 0.05)` associations with $p_{Steiger} < 0.05$).

@@

![](../images/dag-01.png)

`r fig_nums("dag_figure")`

@@

```{r steiger_sensitivity_plot, fig.width=12, fig.height=6 }

@@

`r fig_nums("steiger_sensitivity_plot")`

@@

```{r shakhplot}

`r fig_nums("shakhplot")`

@@

where $\hat{x} = \hat{\alpha}_1 + \hat{\beta}_1 g$. Writing in scale free terms, $\rho_{g, x}$ denotes the correlation between $g$ and the exposure variable $x$, and it is expected that $\rho_{g, x} > \rho_{g, y}$ because $\rho_{g, y} = \rho_{g, x}\rho_{x, y}$, where $\rho_{x, y}$ is the causal association between $x$ and $y$ (which is likely to be less than 1).

In the presence of measurement error in $x$ and $y$, however, the Steiger test will instead be assessing the inequality $\rho_{g, x_o} > \rho_{g, y_o}$. In order to assess how sensitive the Steiger test is to measurement error, we can evaluate the range of potential values of measurement error in $x$ and $y$ over which the empirical result from the Steiger test would return the wrong value.

For different values of $\rho_{x,x_o}$, $\rho_{g,x} = \frac{\rho_{g, x_o}}{\rho_{x,x_o}}$ and $\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1$. For different values of $\rho_{y,y_o}$, $\rho_{g,y} = \frac{\rho_{g, y_o}}{\rho_{y,y_o}}$ and $\rho_{g,y_o} \leq \rho_{y,x_o} \leq 1$.

Where $z = \rho_{g,y} - \rho_{g,x} < 0$ the measurement error range in $x_o$ and $y_o$ will not change the qualitative inference of the direction of causality. To evaluate the sensitivity, $S$, of the Steiger test to measurement error, the objective is to calculate the proportion of the total absolute volume of this function that is above the $z = 0$ plane,

@@

$V_{z}$, the total volume of the function (`r cfig("steiger_sensitivity_plot")`), can be obtained analytically by solving:

@@

The sensitivity, $S$, can take values from 0 to 0.5, where low values suggest there are
